\bf An Examination of Sport Climbing Competition Format and Scoring System

### 2021-08-04 manuscript/Rmd/draft.Rmd
Intro edit

abstract: |
  The purpose of this paper is to investigate the controversial competition format and scoring system of sport climbing, which is one of the sports making its debut on the Olympics stage at Tokyo 2020. Climbing Olympians will participate in three disciplines (speed climbing, bouldering, and lead climbing) for a single set of medals, and a multiplied score of climbers' rankings in all three disciplines will be used to determine the overall standings. This work finds great evidence from historical data and simulations that speed climbing speed climbing should be separated from the combined format and have its own medals. Furthermore, it is shown that the product of ranks scoring system violates the independence of irrelevant alternatives property.

@@

#While there are three main disciplines in sport climbing, the IOC only allotted one set of medals per gender.  So rather than choosing only a single discipline, they chose to combine all three of the events together to create a combined sport climbing event.

@@

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world (CITATION NEEDED). In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint" (CITATION NEEDED). She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

@@

The 8 climbers with the lowest score (i.e. lowest product of ranks across the three disciplines) in the qualification round advance to the finals where they once again compete in all three disciplines. The final score for each person in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round, with the main difference being that there are only 8 competitors in the final as opposed to 20 in the qualification round.  The climbers with the lowest, second lowest, and third lowest product of ranks in the final wins the gold, silver, and bronze medal, respectively.

This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.

>>>>

abstract: |
  The purpose of this paper is to investigate the controversial competition format and scoring system of sport climbing, which is one of the sports making its debut on the Olympics stage at Tokyo 2020. Climbing Olympians will participate in three disciplines: speed climbing, bouldering, and lead climbing for a single set of medals, and a multiplied score of climbers' rankings in all three disciplines will be used to determine the overall standings. This work finds great evidence from historical data and simulations that speed climbing speed climbing should be separated from the combined format and have its own medals. Furthermore, it is shown that the product of ranks scoring system violates the independence of irrelevant alternatives property.

@@

<!-- 
While there are three main disciplines in sport climbing, the IOC only allotted one set of medals per gender.  So rather than choosing only a single discipline, they chose to combine all three of the events together to create a combined sport climbing event. 
-->

@@

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world. In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint". She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

@@

The 8 climbers with the lowest score (i.e. lowest product of ranks across the three disciplines) in the qualification round advance to the finals where they once again compete in all three disciplines. The final score for each person in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round, with the main difference being that there are only 8 competitors in the final as opposed to 20 in the qualification round.  The climbers with the lowest, second lowest, and third lowest product of ranks in the final wins the gold, silver, and bronze medal, respectively. This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.

### 2021-08-04 manuscript/Rmd/draft.tex
Intro edit

\bigskip
\begin{abstract}
The purpose of this paper is to investigate the controversial
competition format and scoring system of sport climbing, which is one of
the sports making its debut on the Olympics stage at Tokyo 2020.
Climbing Olympians will participate in three disciplines (speed
climbing, bouldering, and lead climbing) for a single set of medals, and
a multiplied score of climbers' rankings in all three disciplines will
be used to determine the overall standings. This work finds great
evidence from historical data and simulations that speed climbing speed
climbing should be separated from the combined format and have its own
medals. Furthermore, it is shown that the product of ranks scoring
system violates the independence of irrelevant alternatives property.
\end{abstract}

@@

\#While there are three main disciplines in sport climbing, the IOC only
allotted one set of medals per gender. So rather than choosing only a
single discipline, they chose to combine all three of the events
together to create a combined sport climbing event.

@@

The decision to combine the three climbing events and only award one set
of medals for both men's and women's events in the Olympics has received
a large amount of criticism from climbing athletes all over the world
(CITATION NEEDED). In a series of interviews conducted by Climbing
Magazine in 2016, a number of climbers shared their thoughts and
concerns about the new Olympics climbing format. Climber Lynn Hill
compared the idea of combining speed climbing, bouldering, and lead
climbing to ``asking a middle distance runner to compete in the sprint''
(CITATION NEEDED). She then added ``Speed climbing is a sport within our
sport''. Other climbers also hold the same opinion as Hill regarding
speed climbing, using words and phrases like ``bogus'', ``a bummer'',
``less than ideal'', ``not in support'', and ``cheesy and unfair'' to
describe the new combined competition format. Courtney Woods stated
``Speed climbers will have the biggest disadvantage because their realm
isn't based on difficult movements''. Mike Doyle believed ``Honestly,
the people that will suffer the most are the ones that focus only on
speed climbing. Those skills/abilities don't transfer as well to the
other disciplines''. Most climbers also expressed their hope for a
change in the format in future tournaments, with some calling for each
discipline to have its own set of medals. See \citet{blanchard2016} for
more information on these interviews.

@@

The 8 climbers with the lowest score (i.e.~lowest product of ranks
across the three disciplines) in the qualification round advance to the
finals where they once again compete in all three disciplines. The final
score for each person in the final stage is determined by multiplying
their ranks in each discipline, similar to the qualification round, with
the main difference being that there are only 8 competitors in the final
as opposed to 20 in the qualification round. The climbers with the
lowest, second lowest, and third lowest product of ranks in the final
wins the gold, silver, and bronze medal, respectively.

This type of scoring system heavily rewards high finishes and relatively
ignores poor finishes. For instance, if climber A finished 1st, 20th,
and 20th and climber B finished 10th, 10th, and 10th, climber B would
have a score of 1000 whereas climber A would have a much better score of
400, despite finishing last in 2 out of 3 of the events.

>>>>

\bigskip
\begin{abstract}
The purpose of this paper is to investigate the controversial
competition format and scoring system of sport climbing, which is one of
the sports making its debut on the Olympics stage at Tokyo 2020.
Climbing Olympians will participate in three disciplines: speed
climbing, bouldering, and lead climbing for a single set of medals, and
a multiplied score of climbers' rankings in all three disciplines will
be used to determine the overall standings. This work finds great
evidence from historical data and simulations that speed climbing speed
climbing should be separated from the combined format and have its own
medals. Furthermore, it is shown that the product of ranks scoring
system violates the independence of irrelevant alternatives property.
\end{abstract}

@@

@@

The decision to combine the three climbing events and only award one set
of medals for both men's and women's events in the Olympics has received
a large amount of criticism from climbing athletes all over the world.
In a series of interviews conducted by Climbing Magazine in 2016, a
number of climbers shared their thoughts and concerns about the new
Olympics climbing format. Climber Lynn Hill compared the idea of
combining speed climbing, bouldering, and lead climbing to ``asking a
middle distance runner to compete in the sprint''. She then added
``Speed climbing is a sport within our sport''. Other climbers also hold
the same opinion as Hill regarding speed climbing, using words and
phrases like ``bogus'', ``a bummer'', ``less than ideal'', ``not in
support'', and ``cheesy and unfair'' to describe the new combined
competition format. Courtney Woods stated ``Speed climbers will have the
biggest disadvantage because their realm isn't based on difficult
movements''. Mike Doyle believed ``Honestly, the people that will suffer
the most are the ones that focus only on speed climbing. Those
skills/abilities don't transfer as well to the other disciplines''. Most
climbers also expressed their hope for a change in the format in future
tournaments, with some calling for each discipline to have its own set
of medals. See \citet{blanchard2016} for more information on these
interviews.

@@

The 8 climbers with the lowest score (i.e.~lowest product of ranks
across the three disciplines) in the qualification round advance to the
finals where they once again compete in all three disciplines. The final
score for each person in the final stage is determined by multiplying
their ranks in each discipline, similar to the qualification round, with
the main difference being that there are only 8 competitors in the final
as opposed to 20 in the qualification round. The climbers with the
lowest, second lowest, and third lowest product of ranks in the final
wins the gold, silver, and bronze medal, respectively. This type of
scoring system heavily rewards high finishes and relatively ignores poor
finishes. For instance, if climber A finished 1st, 20th, and 20th and
climber B finished 10th, 10th, and 10th, climber B would have a score of
1000 whereas climber A would have a much better score of 400, despite
finishing last in 2 out of 3 of the events.
